\setupoutput[pdftex]

\setupcolors[state=start]

\usemodule[units]

\setuppagenumbering[location=footer]

\setuppagenumbering[state=start]

\setupwhitespace[big]

\setuplayout[leftmargin=1cm, rightmargin=1cm]
\setuplayout[leftedge=0cm, rightedge=0cm]
\setuplayout[header=1cm, footer=1cm]
\setuplayout[top=0cm, bottom=0cm]

\setupheads[alternative=middle]

\starttext
\title{Fitting Equations in Analysis}

\setupheads[alternative=normal]


\subject{\bf {Protein-Ligand Fast Exchange Kd Equation}}

This equation is for fitting Kd using shifts measured at varying concentrations of ligand (and
sometimes also protein).

We have the reaction $A + B \rightleftharpoons AB$.  We take $A$ to be the protein (or other molecule
that binds) and $B$ to be the ligand.  We vary the amount of $B$ and normally we keep the amount of $A$
fixed.

Let $[A]$ be the concentration of $A$ in the unbound state, $[B]$ be the concentration
of $B$ in the unbound state and $[AB]$ be the concentration of the protein-ligand complex.

Let $a = [A] + [AB]$ be the total concentration of $A$.  Let $b = [B] + [AB]$ be the
total concentration of $B$.

The equilibrium dissociation constant is

$$K_d = {[A][B] \over [AB]}$$

We have

$$\eqalign{[A] &= {K_d [AB] \over [B]} \cr
  &= {K_d (a-[A]) \over [B]} \cr
  &= {K_d (a-[A]) \over b-[AB]} \cr
  &= {K_d (a-[A]) \over b-a+[A]} \cr}$$

Multiply up to get

$$[A] (b-a+[A]) = K_d (a-[A])$$

and so

$$[A]^2 + (b-a+K_d) [A] - aK_d = 0$$

Or in other words

$$\eqalign{[A] &= {-(b-a+K_d) + \sqrt{(b-a+K_d)^2 + 4aK_d} \over 2} \cr
  &= {-(b-a+K_d) + \sqrt{(b+a+K_d)^2 - 4ab} \over 2} \cr}$$

(We cannot have the other root because it would give a negative $[A]$, as can
be seen from the first line above.)  Then

$${a-[A] \over a} = {b+a+K_d - \sqrt{(b+a+K_d)^2 - 4ab} \over 2a}$$

The percentage of bound $A$ is

$$r_{bound} = {[AB] \over a} = {a-[A] \over a}$$

and the percentage of unbound $A$ is

$$r_{free} = 1 - r_{bound}$$

Let $\delta_{bound}$ be the chemical shift of the bound $A$ and $\delta_{free}$ be
the chemical shift of the unbound $A$.

For fast exchange we have that the observed chemical shift is

$$\eqalign{\delta_{obs} &= r_{bound} \delta_{bound} + r_{free} \delta_{free} \cr
  &= r_{bound} \delta_{bound} + (1-r_{bound}) \delta_{free} \cr
  &= \delta_{free} + (\delta_{bound}-\delta_{free}) r_{bound} \cr
  &= \delta_{free} + (\delta_{bound}-\delta_{free}) \left ( {a-[A] \over a} \right ) \cr
  &= \delta_{free} + (\delta_{bound}-\delta_{free}) \left ( {b+a+K_d - \sqrt{(b+a+K_d)^2 - 4ab} \over 2a} \right ) \cr }$$

We normally know what $\delta_{free}$ is and so we introduce

$$\Delta \delta_{obs} = \delta_{obs} - \delta_{free}$$

and similarly

$$\Delta \delta_\infty = \delta_{bound}-\delta_{free}$$

in which case the equation becomes

$$\Delta \delta_{obs} = \Delta \delta_\infty \left ( {b+a+K_d - \sqrt{(b+a+K_d)^2 - 4ab} \over 2a} \right )$$

We know $a$ and $b$.  Sometimes we know $\delta_{bound}$ in which case the only thing left to fit is $K_d$.
Sometimes we do not know $\delta_{bound}$ in which case we have to fit that as well (so equivalently
$\Delta \delta_\infty$).

Let

$$\eqalign{A &= \Delta \delta_\infty / 2 \cr
           B &= 1 + K_d/a \cr
           x &= b/a \cr
           y &= \Delta \delta_{obs} \cr}$$

Then the above equation becomes

$$y = A \left ( B+x - \sqrt{(B+x)^2 - 4x} \right )$$

which is the equation used in Analysis.  We have shift measurements at various $b$.
We fit both $A$ and $B$ and we also use a weighted sum over
the (normally two) dimensions of the shift to make a single scalar.  This is
not optimal.  The ratio $y/A$ is independent of dimension.  In other words,
all dimensions should (or could) provide an estimate of $K_d$, and combining
the shifts together loses information.

Note that

$$K_d = a(B-1)$$

This fitting only makes sense if $b$ is kept constant throughout the experiments.  If $b$ also
varies then we have that $y$ has two dependent variables (in effect, $a$ and $b$).  The Analysis
code would need some changing to cope with that.

A description of the fitting equation is on the web at

$$http://structbio.vanderbilt.edu/chazin/wisdom/kdcalc.htm$$

The correspondence with the formulae in that webpage is:

$$\eqalign{Q_a &= \delta_{obs} \cr
           Q_0 &= \delta_{free} \cr
           Q_{max} &= \delta_{bound} \cr
           [P_{tot}] &= a \cr
           [A_{tot}] &= b \cr}$$

\subject{\bf {Monomer-Dimer Fast Exchange Kd Equation}}

We have the reaction $A + A \rightleftharpoons AA$.  Let 
Let $[A]$ be the concentration of the monomer and $[AA]$
the concentration of the dimer.  Let $a = [A] + 2[AA]$ be the total concentration of $A$.

The equilibrium dissociation constant is

$$K_d = {[A]^2 \over [AA]}$$

Define the ratio

$$r = {[AA] \over a}$$

or in other words

$$[AA] = ra$$

Then we also have

$$[A] = a - 2[AA] = (1-2r)a$$

(Note that $0 \le r \le 1/2$.)  Therefore we have

$$K_d = {(1 - 2r)^2 \over r} a$$

or in other words

$$4r^2 - 4r + 1 = (1 - 2r)^2 = {K_d \over a} r$$

and so

$$r^2 - \left ( 1+{K_d \over 4a} \right ) r + {1 \over 4} = 0$$

And thus

$$\eqalign{r &= {1 \over 2} \left ( 1+{K_d \over 4a} - \sqrt{\left (1 + {K_d \over 4a} \right )^2 - 1} \right ) \cr
  &= {1 \over 8a} \left ( K_d + 4a - \sqrt{\left (K_d + 4a \right )^2 - 16a^2} \right ) \cr}$$

The positive root cannot be taken because that would make $r > 1/2$.

Let $\delta_A$ be the chemical shift of the monomer and $\delta_{AA}$ be
the chemical shift of the dimer, and define $\Delta \delta_\infty = \delta_{AA} - \delta_A$.
In general we do not know either $\delta_A$ or $\delta_{AA}$.

For fast exchange we have that the observed chemical shift is

$$\eqalign{\delta_{obs} &= {[A] \delta_A + 2[AA] \delta_{AA} \over a} \cr
  &= (1-2r) \delta_A + 2r \delta_{AA} \cr
  &= \delta_A + 2r (\delta_{AA} - \delta_A) \cr
  &= \delta_A + 2r \Delta \delta_\infty \cr
  &= \delta_A + {\Delta \delta_\infty \over 4a} \left ( K_d + 4a - \sqrt{\left (K_d + 4a \right )^2 - 16a^2} \right ) \cr}$$

There is no point subtracting anything here since we do not know either $\delta_A$ or $\delta_{AA}$
in general.

Let

$$\eqalign{A &= \Delta \delta_\infty \cr
           B &= K_d \cr
           C &= \delta_A \cr
           x &= a \cr
           y &= \delta_{obs} \cr}$$

Then the above equation becomes

$$\eqalign{y &= A \left ( B+4x - \sqrt{(B+4x)^2 - 16x^2} \right ) / 4x + C \cr
             &= A \left ( 1 + B/4x - \sqrt{(1+B/4x)^2 - 1} \right ) + C \cr}$$

which is the equation used in Analysis.

\stoptext
